---
title: "Overview of R -- Part I"
author: "Asad Haris"
date: "12 September, 2019"
output:
extra_dependencies: xcolor
---

## Overview of Lecture

Today we will (roughly) cover the following topics: 

1. How to create an object
2. Operators
3. How to generate data &  Data types
4. ~~Introduction to ggformula and mosaic packages~~ (Next Lecture)
5. ~~Coding style~~ (Instead please see excellent [guide by Hadley Wickham](https://style.tidyverse.org/))

## Review basic computations

- Recall basic computations from last lecture

```{r, collapse=TRUE}
2+3
log(4)  # Natural log
sin(2)  # Sin function
2e+3  # Equal to 2 * 10^3
8^(1/3)
sqrt(4)
```

- The symbol `#` is used to add comments to code

## Creating objects

- We can use `<-`, `->` & `=`

```{r, collapse=TRUE}
# Using <-, -> & =
num.1 <- 2
3 -> num.2
num.3 = 8

# Case-Sensitive
Num.1 <- 20

# Invalid options, Cannot start with number
# 45_23 <- 13
# 4num <- 4

# Valid to use Alpha+numeric
num4 <- 4

# Can assign multiple objects at the same time
x <- y <- z <- 10

# Old values get replaced
num.1 <- 33 
```

- Do NOT assign the single letter names c, g, t, C, D, F, I and T as they are default names that are used by R. For instance, T and F are abbreviations for TRUE and FALSE in logical operations. We should avoid using names that are already used by the system. 

- There is no limit on length of variable names. Using appropriate or informative variable names can be helpful for reading code and reproducibility. 

```{r,collapse=TRUE}
# calculations with variables
num.students <- 30
num.male <- 13

num.female <- num.students - num.male
prec.male <- num.male/num.students * 100
prec.female <- num.female/num.students * 100
```

- We can view objects in the environment window, or we could use the `ls()` function.
- Objects can be deleted using `rm()`

```{r,collapse=TRUE}
ls()
rm(num.students)

# Deleting all objects in a session
list.of.objects <- ls()
rm(list = list.of.objects)
```

## Data types and structures

**Firstly, a side-note about functions:**

    - Functions are `R` commands which take in inputs and return an output, the general template will be `some.function(input1, input2, input3)`. 
    - All plot, data summary, data analysis tools in R are functions. 
    - Apart from google, your main best friend should be the function help-page:
        + `?function-name`


### Vectors (and data types)

- R is a vectorized language, operations/functions can be applied to a whole vector. 
- The basic way to define a vector is using the `c()` function, the 'c' stands for concatenate.
- We use vectors to discuss the various data-types

```{r,collapse=TRUE}
# A numeric vector
age <- c(22, 21, 23, 22, 24, 21, 21)
class(age)

# A character vector
class.names <- c("Robert", "Alice", "Mark", "Kate",
              "Charlie", "Victor", "John")
class(class.names)

# A logical vector
EPIB.student <- c(TRUE, TRUE, TRUE, FALSE, FALSE, 
                  TRUE, TRUE)
class(EPIB.student)
```

- Vectors can be named, we will later see almost all objects can be named
- Names can be assigned manually or using an existing vector of names
```{r,collapse=TRUE}
names(age) <- class.names
age

grades <- c("Robert" = "A+", "Alice" = "A+", "Mark" = "B+",
            "Kate" = "B-", "Charlie" = "C+", "Victor" = "A-", "John" = "B+")
print(grades)
```

#### Generating data: vectors
- We have already seen data generation using `c()` function
- Other functions for generating vectors:
    + `seq()`
    + `rep()`
    + `sequence()`
    + Random numbers*: `rnorm`, `rbinom`, `runif`, `sample` etc. 
    
```{r,collapse=TRUE}
c(-1, 5.44, 100, 34123)

-1:10 # By increments of 1.

seq(from = 0.33, to = 9.33, by = 3)

seq(from = 0, to = 1, length = 5)

rep(1.2, times = 5)

rep(c("six", "one", "three"), each = 2)

c(6, 1, 3, rep(seq(from = 3, to = 5, by = 0.5), times = 2))

sequence(5)

sequence(c(6, 1, 3))

```

- We can generate a random sample from given data using `sample`
  
```{r,collapse=TRUE}
#In order to be able to replicate random sampling results, we should set a seed
set.seed(2019)

# Sample with replacement
sample(1:30, size = 10, replace = TRUE)

# Sample without replacement
sample(1:30, size = 10, replace = FALSE)

# Sample names
sample(class.names, size = 3)
```

- R has built in functions to sample data from given distributions
```{r,collapse=TRUE}
# Uniform random variables from [-1, 1]
runif(n = 10, min = -1, max = 1)

# Gaussian/Normal random variables with mean = 0, sd = 1
rnorm(n = 10, mean = 0, sd = 1)

# Binomial(5, 0.3) distribution
rbinom(n = 10, size = 5, prob = 0.3)
```


### Matrix

- Most data structures are made from the basic vectors

```{r,collapse=TRUE}
mymatrix1 <- matrix(c(1:12), nrow = 4, ncol = 3, byrow = TRUE)
print(mymatrix1)

mymatrix2 <- matrix(c(1:12), nrow = 4, ncol = 3, byrow = FALSE)
print(mymatrix2)

rownames <- c("Subject A", "Subject B", "Subject C", "Subject D")
colnames <- c("Sample 1", "Sample 2", "Sample 3")
rownames(mymatrix1) <- rownames
colnames(mymatrix1) <- colnames
print(mymatrix1)

```

- We have to be careful when creating matrices and other complex structures
```{r,collapse=TRUE}
# Sometimes R gives a warning
bad.matrix1 <- matrix(c(1:11), nrow = 4, ncol = 3, byrow = TRUE)
print(bad.matrix1)

# Sometimes there is no warning, be careful!
bad.matrix2 <- matrix(c(1:4), nrow = 6, ncol = 6)
print(bad.matrix2)

```

### Arrays

- Mathematically, scalars, vectors and matrices are all arrays of different dimensions
    + Scalar: 1 x 1 array
    + Vector of length k: 1 x k array
    + Matrix of dimension m x n: m x n array

- R treats every array below 3 dimensions differently but they are essentially not very different. 

- Now let's look at a 3-dimensional array.
```{r,collapse=TRUE}
myarray <- array(1:24, dim = c(4,3,2))
print(myarray)
```

- Demonstrating Arrays using fake data
    + Disease: 1 (Yes) and 0 (No)
    + Drug: 1 (Exposed) and 0 (Unexposed)
    + BMI category: 1, 2, 3
    + Age category: 1, 2, 3, 4

```{r,collapse=TRUE}
set.seed(2019) 

# Randomly assign ~20% of patients to have disease.
disease <- sample(c(0,1), size = 100, replace = TRUE, prob = c(0.2, 0.8))

# Randomly assign ~40% of patients to take drug.
drug <- sample(c(0,1), size = 100, replace = TRUE, prob = c(0.4, 0.6))

bmi.cat <- sample(1:3, size = 100, replace = TRUE) # Randomly assign BMI categories
age.cat <- sample(1:4, size = 100, replace = TRUE) # Randomly assign age categories

# Don't worry about data.frame : coming up next!
data <- data.frame(drug, disease, bmi.cat, age.cat)
head(data)

# The table below shows the first 6 rows of the fake dataset.
# This is a typical dataset you will see in Epidemiology.
# Each row is a patient, with their own information.
# Goal is to assess the association between disease and drug (drug safety).

# By tabulating the data, we can assess the association (EPIB 601 material).
# If we only tabulate drug and disease, we get a 2x2 table, which is a matrix or a 2-dimensional array.
# 1st dimension: drug, 2nd dimension: disease
table(data[c("drug","disease")])
prop.table(table(data[c("drug","disease")]), margin = 1)

# This may not be enough, we want to see how people with different BMI may differ (confounder, also 601 material).
# We now need a 2x2x3 table, which is a 3-dimensional array.
# 1st dimension: drug, 2nd dimension: disease, 3rd dimension: bmi.cat
table(data[c("drug", "disease", "bmi.cat")])

# Further include age to see how age category comes into the association
# We now need a 2x2x3x4 table, which is a 4-dimensional array.
# 1st dimension: drug, 2nd dimension: disease, 3rd dimension: bmi.cat, 4th dimension: age.cat

#table(data)

```


### Data frames

Data frame is the most commonly used member of the data types family in R. A data frame is a generalization of a matrix, in which different columns may have different modes. All elements of any column must have the same mode, i.e. all numeric or all factor, or all character.

```{r, collapse=TRUE}

dat.1 <- data.frame(class.names, age, grades, EPIB.student)
print(dat.1)

str(dat.1) # checking the structure of an object

```

- You might have seen the `NA` or `NaN` symbols in R. 
    + NA can b used for 

```{r,collapse=TRUE}
0/0

# We can also assign NaN values to data
dat.1$age <- c(22, 21, 23, 22, 24, NA, NA)
```



### Lists: special types of vectors

- Lists are like vectors, only the objects can be of different types
    + We can have a list of numbers or characters
    + We can have a list of some numbers, some characters and some logical

- Lists can also hold other data structures
    + We can have a list where each element is a vector
    + or a matrix
    + We could even have a list of lists!

```{r,collapse=TRUE}

mylist <- list("Red", factor(c("a","b")), c(21,32,11), TRUE)
print(mylist)

str(mylist)

```

### Factors

- A lot like character vectors
    + They have an additional `levels` part to them
    + This type of data structure is useful with other R functions (later)

```{r,collapse=TRUE}
ch.letter <- rep(letters[1:3], each = 4)
print(ch.letter)

class(ch.letter)

fac.letter <- rep(as.factor(letters[1:3]), each = 4)
print(fac.letter)
# Note the additional 'Levels: a b c' in the output

class(fac.letter)

levels(fac.letter) <- c("Alpha", "Bravo", "Charlie")
print(fac.letter)

```


### Conversions between different types:

Conversion between these data structures
    - as.vector
    - as.matrix
    - as.array
    - as.data.frame
    - ...
Check whether your R object has certain data structure
    - is.vector
    - is.matrix
    - is.array
    - is.data.frame
    - ...
    
## More on Operations

So far we have seen simple operations on scalars. R allows us to do much more with vectors and matrices.

### Numerical operations

```{r,collapse=TRUE}
a <- c(1, 8, 8)
b <- c(2, 8, 4)

a+1 # here 1 is considered as a vector (1, 1, 1)

a+b

a*b

a^2

c <- matrix(c(1,2,3,4), nrow = 2, byrow = T)
d <- matrix(c(5,6,7,8), nrow = 2, byrow = F)
print(c); print(d)

c+1

c+d

c*d

c^2

a %*% b

a %o% b

c %*% d

c; t(c)

diag(c)

det(c)

```

### Logical operators

```{r, collapse=TRUE}
# Recall vector a and b from above.
print(a); print(b)

a == b # Equal or not?

a != b # Not equal?

a > b

a <= b

# And
a; b
a>5 & b>5

# Or
a>=5 | b>=5

"ABC" == "ABC"

"ABC" == "abc"

TRUE + TRUE + FALSE # True = 1, False = 0.
```

## Data viziulization 

We will begin with some basic data visualization using the `ggformula` package. 

```{r,message=FALSE,warning=FALSE}
library(ggformula)
```


Basic idea for ggformula is as follows: 

<center> <h3>FunctionName (Outcome Variable ~ Explanatory Variable(s), data = myData)</h3> </center>

We will use a number of existing dataframes in R for this demonstration. To see list of available data sets in R, use the command `data()`.

We will consider the following:
    - Basic Uni variate plots: histograms, density, boxplots
    - Basic bivariate plots: boxplots, scatter, line, smoothing line
    - Other features

### Histograms & density plots

```{r,collapse=TRUE}
# First consider the USArrests Data.
# Samples (n) = 50, and 4 variables:
# 1. Number of Murder arrests /100,000, 
# 2. Number of Assault arrests /100,000
# 3. Percent of urban pop, 
# 4. Number of rape arrests /100,000 

ggformula::gf_histogram(~Murder, data = USArrests)

# Histogram with some options
ggformula::gf_histogram(~Murder, data = USArrests, bins = 20,
                        fill = "green", color = "red", 
                        title = "Murder Arrests in USA",
                        xlab = "Number of Murder Arrests/100,000")

# Using base-R
hist(USArrests$Murder)
# With options
hist(USArrests$Murder, col = "red", xlab = "Number of Murder Arrests/100,000", main = "Murder Arrests in USA", breaks = 20)
```


### Box plots

```{r}
# Simple boxplot using ggformula and base-R
ggformula::gf_boxplot(~Murder, data = USArrests, 
                      color = "red", fill = "green") 

boxplot(USArrests$Murder, 
        xlab = "",
        ylab = "Number of Arrests/100,000", col = 2)

# Easy to do multiple boxplots in base-R
boxplot(list(USArrests$Murder, USArrests$Assault, USArrests$Rape), 
        names = c("Murder", "Assault", "Rape"), notch = TRUE,
        ylab = "Number of Arrests/100,000", col = 2:4,
        log = "y") # log = "y", log-scales the y-axis.

```


### Continuous vs. Categorical: boxplots

- Boxplots can be made by a categorical variable
- We look at the `esoph` data now:
    + `agegp`, `alcgp` and `tobgp`: 3 categorical variables for age, alcohol consumption and tobacco consumption.
    + `ncases`: number of cases of esophageal cancer
    + `ncontrols`: number of controls

```{r}
# No. of cases by age group
ggformula::gf_boxplot(ncases~agegp, data = esoph, fill = "green",
                      xlab = "Age Group (years)", ylab = "No. of Cases")
boxplot(ncases ~ agegp, data = esoph, col = "lightgreen",
        xlab = "Age Group (years)", ylab = "# of Cases",
        outline = FALSE)
```

#### Faceting: a very useful feature

- Faceting is a useful feature when we might have multiple variable we wish to group by
- An easy feature for `ggformula` and `ggplot` but harder to do in base-R.

```{r}
ggformula::gf_boxplot(ncases ~ agegp | alcgp, 
                      data = esoph, fill = ~alcgp,
                      xlab = "Age Group (years)", ylab = "No. of Cases")

par(mfrow = c(2,2))
boxplot(ncases ~ agegp, data = esoph, subset = (alcgp == "0-39g/day"),
        col = 2)
boxplot(ncases ~ agegp, data = esoph, subset = (alcgp == "40-79"),
        col = 3)
boxplot(ncases ~ agegp, data = esoph, subset = (alcgp == "80-119"),
        col = 4)
boxplot(ncases ~ agegp, data = esoph, subset = (alcgp == "120+"),
        col = 5)
par(mfrow = c(1,1))
```
### Continuous vs. Categorical: barplots

- Barplots are for known heights which need to plotted. 
    + We will use the `Titanic` dataset, see `?Titanic`
    + We will first convert it to a `data.frame`
    
```{r}
dat <- as.data.frame(Titanic)

ggformula::gf_col(Freq ~ Survived|Class, 
                  data = subset(dat,Age == "Adult" & Sex == "Male"),
                  fill = ~ Class)

# Frequency of Adult, Men who survived by class
barplot(Freq ~ Survived + Class, data = dat,
        subset = Age == "Adult" & Sex == "Male", legend = TRUE,
        col = c("lightgreen", "cyan") )
```

#### Adding layers in ggformula

- In `ggformula`, each plot is a layer and we can stack the layers together
    + E.g. we could have a scatter plot and add a layer of loess smoothers
    + Or (example below), we could add error-bars or confidence intervals to a plot
- Adding layers in ggplot is done using the `%>%` symbol
```{r}
# We consider the hypothetical scenario where there might be measurement
# error in counting the people. 
# Assume we know the SE, we can then make 95% Wald confidence intervals
# as: (est - 1.96*SE, est + 1.96*SE)

# Generate 95% confidence interval with SE = 2
dat$lower.CI <- dat$Freq - 1.96*10
dat$upper.CI <- dat$Freq + 1.96*10

ggformula::gf_col(Freq ~ Survived|Class, 
                  data = subset(dat,Age == "Adult" & Sex == "Male"),
                  fill = ~ Class) %>%
  ggformula::gf_errorbar(lower.CI + upper.CI ~ Survived|Class, width = 0.2)

```


### Continuous vs Continous (Coming up next)
